---
title: "Yellow Taxi Trip Data Analysis"
subtitle: "NYC | Taxi & Limousine Commision"
date: "2025-06-07"
author: "Jaka Ahlin"
format: 
    html:
      toc: true
      toc-location: left
      toc-depth: 4
      toc-float: true
      embed-resources: true
editor: visual
execute:
    warning: false
    message: false
    error: true
---

# Introduction

Using the Yellow Taxi Trip Data I will analyze how different aspects of the trip affect the tip amount. Let's see if the day of the week, trip distance, pickup and dropoff locations, dropoff hour, payment type, vendor, number of passengers and other aspects influence the tips given to drivers.

# Presentation of the Data

## Load packages

```{r}
library("arrow")
library("gt")
library("janitor")
library("psych")
library("skimr")
library("dplyr")
library("tidyverse")
library("ggplot2")
library("DT")
```

## Read data

```{r}
df <- read_parquet("../archive/yellow_tripdata_2025-01.parquet")
```

## Clean names

```{r}
df <- clean_names(df)
```

## View data

```{r}
df %>% 
  head() %>% 
  gt()
```

```{r}
skim(df)
```

## Column descriptions

| Field Name | Description |
|------------------------------------|------------------------------------|
| VendorID | A code indicating the TPEP provider that provided the record.<br>1 = Creative Mobile Technologies, LLC<br>2 = Curb Mobility, LLC<br>6 = Myle Technologies Inc<br>7 = Helix |
| tpep_pickup_datetime | The date and time when the meter was engaged. |
| tpep_dropoff_datetime | The date and time when the meter was disengaged. |
| passenger_count | The number of passengers in the vehicle. |
| trip_distance | The elapsed trip distance in miles reported by the taximeter. |
| RatecodeID | The final rate code in effect at the end of the trip.<br>1 = Standard rate<br>2 = JFK<br>3 = Newark<br>4 = Nassau or Westchester<br>5 = Negotiated fare<br>6 = Group ride<br>99 = Null/unknown |
| store_and_fwd_flag | This flag indicates whether the trip record was held in vehicle memory before sending to the vendor, aka "store and forward," because the vehicle did not have a connection to the server.<br>Y = store and forward trip<br>N = not a store and forward trip |
| PULocationID | TLC Taxi Zone in which the taximeter was engaged. |
| DOLocationID | TLC Taxi Zone in which the taximeter was disengaged. |
| payment_type | A numeric code signifying how the passenger paid for the trip.<br>0 = Flex Fare trip<br>1 = Credit card<br>2 = Cash<br>3 = No charge<br>4 = Dispute<br>5 = Unknown<br>6 = Voided trip |
| fare_amount | The time-and-distance fare calculated by the meter. For additional information on the following columns, see https://www.nyc.gov/site/tlc/passengers/taxi-fare.page |
| extra | Miscellaneous extras and surcharges. |
| mta_tax | Tax that is automatically triggered based on the metered rate in use. |
| tip_amount | Tip amount – This field is automatically populated for credit card tips. Cash tips are not included. |
| tolls_amount | Total amount of all tolls paid in trip. |
| improvement_surcharge | Improvement surcharge assessed trips at the flag drop. The improvement surcharge began being levied in 2015. |
| total_amount | The total amount charged to passengers. Does not include cash tips. |
| congestion_surcharge | Total amount collected in trip for NYS congestion surcharge. |
| airport_fee | For pick up only at LaGuardia and John F. Kennedy Airports. |
| cbd_congestion_fee | Per-trip charge for MTA's Congestion Relief Zone starting Jan. 5, 2025. |

# Transformation of the Data

## Adding Vendor Names based on Vendor Id

Let's add vendor names based on the vendor id. These names will be used later.

```{r}
df <- df %>%
  mutate(vendor_name = case_when(
    vendor_id == 1 ~ "Creative Mobile Technologies, LLC",
    vendor_id == 2 ~ "Curb Mobility, LLC",
    vendor_id == 6 ~ "Myle Technologies Inc",
    vendor_id == 7 ~ "Helix"
  ))

df$vendor_name %>% head()
```

## Adding A Payment Type Column

Let's add payment type name based on the payment type id. These names will be used later.

```{r}
df <- df %>%
  mutate(payment_type_name = case_when(
    payment_type == 0 ~ "Flex Fare trip",
    payment_type == 1 ~ "Credit card",
    payment_type == 2 ~ "Cash",
    payment_type == 3 ~ "No charge",
    payment_type == 4 ~ "Dispute",
    payment_type == 5 ~ "Unknown",
    payment_type == 6 ~ "Voided trip"
  ))

df$payment_type_name %>% head()
```

## Date and Time transformation

Transforming the time from POSIXct to date-time format.

```{r}
df <- df %>%
  mutate(across(c(tpep_pickup_datetime, tpep_dropoff_datetime), ymd_hms))

df %>%
  select(tpep_pickup_datetime, tpep_dropoff_datetime) %>%
  head()
```

## Adding a day in the week column

It's important to know the day of the week the trip was made to better understand the data and later it will help to analyze the tips.

```{r}
df <- df %>%
  mutate(
    pickup_day = wday(tpep_pickup_datetime, week_start = 1),
    dropoff_day = wday(tpep_dropoff_datetime, week_start = 1)
  )

df %>% 
  select(pickup_day, dropoff_day) %>%
  head()
```

## How many trips were during the weekend (Saturday and Sunday) and during the weekdays (Monday - Friday) percentage wise?

In the table below is shown the percentage of trips made during the weekdays and weekends.

```{r}
df %>%
  filter(pickup_day %in% 1:5 | dropoff_day %in% 1:5) %>%
  summarise(Mon_Fri=n() / nrow(df) * 100, Sat_Sun = 100 - Mon_Fri)
```

## Do drivers get more tips during the weekday or weekend?

Surprisingly, drivers got more tips during the weekdays rather than the weekends, which can be seen from the average tip amount in the table below.

```{r}
df %>%
  mutate(day_type = ifelse(pickup_day %in% 1:5 & dropoff_day %in% 1:5, "Weekday", "Weekend")) %>%
  group_by(day_type) %>%
  summarise(avg_tip = mean(tip_amount))
```

## Do the tips differ between the vendors?

There are four different vendors in the dataset with two of them having the most amount of trips and also similar average tip amounts. While Helix has the highest average tip, it has a small amount of trips compared to the other vendors, which could skew the results. Similar problem we can see with the Myle Technologies Inc vendor.

```{r}
df %>%
  group_by(vendor_name) %>%
  summarise(avg_tip = mean(tip_amount), num_trips = n()) %>%
  arrange(desc(avg_tip))
```

## Does the payment type affect the tip amount?

The tip amount only gets automatically populated for credit card payments. Cash tips are not included.

```{r}
df %>%
  group_by(payment_type_name) %>%
  summarise(avg_tip = mean(tip_amount), num_trips = n()) %>%
  arrange(desc(avg_tip))
```

## Do the tips differ based on the pickup and dropoff locations?

Trips starting or ending in some locations tend to have higher average tips than others, which can be seen from the table below. It is also important to note the number of trips in each location, as some locations have a very small number of trips, which can skew the results.

```{r}
df %>%
  group_by(pu_location_id) %>%
  summarise(avg_tip = mean(tip_amount), num_trips = n()) %>%
  arrange(desc(avg_tip)) %>%
  datatable()

df %>%
  group_by(do_location_id) %>%
  summarise(avg_tip = mean(tip_amount), num_trips = n()) %>%
  arrange(desc(avg_tip)) %>%
  datatable()

```

# Exploratory Data Analysis

## Do the tips differ based on the number of passengers?

The table is included to show the number of trips for each passenger count, from which we can see that for 7, 8 or 9 passengers there are very few trips, which can skew the results.

```{r}
df %>%
  filter(!is.na(passenger_count)) %>%
  group_by(passenger_count) %>%
  summarise(avg_tip = mean(tip_amount), num_trips = n())

df %>%
  filter(!is.na(passenger_count)) %>%
  group_by(passenger_count) %>%
  summarise(avg_tip = mean(tip_amount)) %>%
  ggplot(aes(x = factor(passenger_count), y = avg_tip, group = 1)) +
  geom_area(fill = "gold") +
  labs(
    title = "Average Tip by Number of Passengers",
    x = "Number of Passengers",
    y = "Average Tip Amount"
  )
```

## Does the trip distance affect the tip amount?

The average tip amount increases with the trip distance, but the number of trips decreases as the trip distance increases, which can skew the results in the second half of the graph.

```{r}
df %>%
  mutate(trip_distance = ceiling(trip_distance)) %>%
  group_by(trip_distance) %>%
  summarise(avg_tip = mean(tip_amount), num_trips = n()) %>%
  filter(num_trips > 1) %>%
  ggplot(aes(x = factor(trip_distance), y = avg_tip)) +
  geom_col(fill = "yellowgreen") +
  labs(title = "Average Tip Amount by Trip Distance",
       x = "Trip Distance (miles)",
       y = "Average Tip Amount") +
  scale_x_discrete(breaks = seq(0, max(df$trip_distance), by = 5))
```

## Does the dropoff hour affect the tip amount?

This line chart nicely presents us the average tip amount by the dropoff hour. We can see that the average tip amount is highest during the late evening hours.

```{r}
df %>%
  mutate(dropoff_hour = hour(tpep_dropoff_datetime)) %>%
  filter(!is.na(dropoff_hour)) %>%
  group_by(dropoff_hour) %>%
  summarise(avg_tip = mean(tip_amount), num_trips = n()) %>%
  ggplot(aes(x = dropoff_hour, y = avg_tip)) +
  geom_line(color = "magenta", size = 1.5) +
  geom_point() +
  scale_x_continuous(breaks = 0:23) +
  labs(
    x = "Dropoff Hour",
    y = "Average Tip Amount",
    title = "Average Tip by Hour"
  )
```

# Interpretation and Conclusion

Given the information presented we can conclude that the tips given to drivers are influenced by various factors like trip distance, dropoff hour, pickup and dropoff locations, weekday or weekend.

# References
- [Yellow Taxi Trip Data](https://www.kaggle.com/datasets/ranjanrakesh51/yellow-taxi-trip-data)
- [R-statistics](https://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html)

